---
title: "Heart Disease Prediction: Model Comparison and Analysis"
author: "James Obanla"
date: "2024-12-31"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1. Introduction: 

One of the reasons for the human population's loss of life is heart disease. This project is focused on comparing models built using exogenous variables such as age, sex, number of major vessels, serum cholesterol, resting blood pressure, chest pain, etc. to predict the target—the presence or absence of heart disease as the dependent variable. The sample size of datasets available for this task is 303 observations and 14 variables containing qualitative and quantitative data. The results derived from the support vector machine and logistic regression models are thoroughly analyzed and compared, leading to data-driven recommendations based on their respective performances
 
### 2. Loading Necessary Packages

Packages for data manipulation, visualization, and machine learning are loaded.

```{r , warning=FALSE, message=FALSE, echo=TRUE}
library(tidyverse)
library(reshape2)
library(ggplot2)
library(caTools)
library(e1071)
library(randomForest)
```


```{r , warning=FALSE, message=FALSE, echo=TRUE}
library(gt)

```


### 3. Loading and Inspecting Data

The heart dataset is loaded, and its structure and initial rows are reviewed.

```{r , warning=FALSE }

data<-read.csv(file.choose(), header = TRUE)

attach(data)

head(data)
```


### 4. Data Cleaning

Handle duplicates, missing values, and encode categorical variables for analysis.

```{r , warning=FALSE}

data %>% filter(duplicated(.))

any(is.na(data))

data <- data %>% drop_na()  ## removing of missing data

data_n <- data %>%
  mutate(
    ChestPain = ifelse(ChestPain == "typical", 1, 
                       ifelse(ChestPain == "asymptomatic", 2, 
                              ifelse(ChestPain == "nonanginal", 3, 
                                     ifelse(ChestPain == "nontypical", 4, NA)))),
    Thal = ifelse(Thal == "normal", 1, 
                  ifelse(Thal == "fixed", 2, 
                         ifelse(Thal == "reversable", 3, NA))),
    Thal = ifelse(is.na(Thal), mean(Thal, na.rm = TRUE), Thal)
  )

tail(data_n) 

summary(data_n)

```


### 5. Exploratory Data Analysis

Boxplots are created to explore variable distributions by the target class.

```{r , warning=FALSE}

# Ensure Target is a factor with the correct levels
data_n$Target <- factor(ifelse(data_n$Target == 0, "no", "yes"), levels = c("no", "yes"))

# Create the boxplot
data1 <- ggplot(data_n, aes(x = Target, y = Age, fill = Target)) +
  geom_boxplot(alpha = 0.8, color = "black") +
  theme_minimal() +
  labs(title = "Boxplot of Age by Target",
       x = "Target",
       y = "Age") +
  scale_fill_manual(values = c("no" = "orange", "yes" = "brown"))

# Display the plot
data1

# Create the boxplot for RestBP by Target
data2 <- ggplot(data_n, aes(x = Target, y = RestBP, fill = Target)) +
  geom_boxplot(alpha = 0.8, color = "black") +
  theme_minimal() +
  labs(title = "Boxplot of RestBP by Target",
       x = "Target",
       y = "Resting Blood Pressure (RestBP)") +
  scale_fill_manual(values = c("no" = "blue", "yes" = "red"))

# Display the plot
data2

# Create the boxplot for Chol by Target
data3 <- ggplot(data_n, aes(x = Target, y = Chol, fill = Target)) +
  geom_boxplot(alpha = 0.8, color = "black") +
  theme_minimal() +
  labs(title = "Boxplot of Chol by Target",
       x = "Target",
       y = "Cholesterol (Chol)") +
  scale_fill_manual(values = c("no" = "purple", "yes" = "yellow"))

# Display the plot
data3

data4 <- ggplot(data_n, aes(x = Target, y = MaxHR, fill = Target)) +
  geom_boxplot(alpha = 0.8, color = "black") +
  theme_minimal() +
  labs(title = "Boxplot of MaxHR by Target",
       x = "Target",
       y = "Maximum Heart Rate Achieved (MaxHR)") +
  scale_fill_manual(values = c("no" = "green", "yes" = "gold"))

data4

data5 <- ggplot(data_n, aes(x = Target, y = Oldpeak, fill = Target)) +
  geom_boxplot(alpha = 0.8, color = "black") +
  theme_minimal() +
  labs(title = "Boxplot of Oldpeak by Target",
       x = "Target",
       y = "ST Depression Induced by Exercise Relative to Rest(Oldpeak)") +
  scale_fill_manual(values = c("no" = "orchid", "yes" = "turquoise"))

data5

```


### 6. Correlation Heatmap

A correlation matrix heatmap is created for independent variables, separated by dependent variable-Target.

```{r , warning=FALSE}
# Select numeric variables for correlation
data6 <- data_n[, c("Age", "Sex", "ChestPain", "RestBP", "Chol", "Fbs", 
                    "RestECG", "ExAng", "Slope", "Ca", "Thal", "MaxHR", "Oldpeak")]

# Ensure Target is treated as a categorical variable
data_n$Target <- factor(data_n$Target)

# Compute correlation matrices separately for each Target category
cor_by_target <- lapply(split(data6, data_n$Target), function(subset) {
  cor(subset, use = "complete.obs")
})

# Convert the correlation matrices into long format and add Target labels
cor_melted_list <- lapply(names(cor_by_target), function(target) {
  melted <- reshape2::melt(cor_by_target[[target]])
  melted$Target <- target
  return(melted)
})

# Combine all melted data into a single data frame
data7 <- do.call(rbind, cor_melted_list)

# Plot the heatmap with correlation values
ggplot(data7, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  geom_text(aes(label = round(value, 2)), size = 4, color = "black") +  # Add correlation values
  scale_fill_gradient(low = "white", high = "blue") +
  facet_wrap(~ Target) +  # Separate heatmaps for each Target category
  labs(title = "Correlation Heatmap by Target",
       x = "Variable 1", y = "Variable 2") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

### 7. Logistic Regression Model

A Logistic Regression model is trained and evaluated.

```{r , warning=FALSE}

# Set a random see so your "random" results are the same as this notebook
set.seed(101) 

# Split up the sample, basically randomly assigns a booleans to a new column "sample"
sample <- sample.split(data_n$Target, SplitRatio = 0.80) # SplitRatio = percent of sample==TRUE

# Training Data
train <-subset(data_n, sample == TRUE)

# Testing Data
test <- subset(data_n, sample == FALSE)


#Fitting Generalized Linear Models

model_lg = glm(Target ~ ., family = binomial(logit), data = train)

summary(model_lg)

# Choose a model by AIC in a Stepwise Algorithm

new.step.model <- step(model_lg)

summary(new.step.model)


```

Creating a confusion matrix using the predict function with type='response' as an argument inside of that function.

```{r, warning=FALSE}

test$predicted.Target = predict(model_lg, newdata=test, type="response")

# Confusion matrix
table(test$Target, test$predicted.Target > 0.5)


```

Confusion matrix values

```{r , warning=FALSE}

TP <- 18  # True Positives
FP <- 4   # False Positives
TN <- 29  # True Negatives
FN <- 10  # False Negatives

# Calculate metrics

accuracy <- (TP + TN) / (TP + FP + TN + FN)

precision <- TP / (TP + FP)

recall <- TP / (TP + FN)

f1_score <- 2 * (precision * recall) / (precision + recall)

specificity <- TN / (TN + FP)

# Print results
cat("Accuracy:", round(accuracy, 4), "\n")

cat("Precision:", round(precision, 4), "\n")

cat("Recall (Sensitivity):", round(recall, 4), "\n")

cat("F1 Score:", round(f1_score, 4), "\n")

cat("Specificity:", round(specificity, 4), "\n")

```

### Interpretation:
 
- Accuracy (78.57%): Indicates that ~79% of predictions were correct.

- Precision (81.82%): Suggests the model effectively avoids False Positives.

- Recall (64.29%): Shows the model misses some True Positives (False Negatives present).

- F1 Score (72.00%): Balances precision and recall, indicating moderate performance.

- Specificity (87.88%): High specificity suggests the model is good at correctly  

identifying negatives ("No").


### 8. Support Vector Machine Model

A Support Vector Machine (SVM) model is trained and evaluated.

```{r , warning=FALSE}

library(e1071)

model_svm <- svm(Target ~ .,data=train)

summary(model_svm)

predicted.values <- predict(model_svm,test[1:13])

```


```{r , warning=FALSE}
# Confusion Matrix 

table(predicted.values,test$Target)

```


```{r , warning=FALSE}

### Confusion matrix vlues

TP <- 20  # True Positives

FP <- 8   # False Positives

TN <- 31  # True Negatives

FN <- 2   # False Negatives


### Calculate metrics

accuracy <- (TP + TN) / (TP + FP + TN + FN)

precision <- TP / (TP + FP)

recall <- TP / (TP + FN)

f1_score <- 2 * (precision * recall) / (precision + recall)

# Print results

cat("Accuracy:", round(accuracy, 4), "\n")

cat("Precision:", round(precision, 4), "\n")

cat("Recall:", round(recall, 4), "\n")

cat("F1 Score:", round(f1_score, 4), "\n")
```

#### Interpretation:

- High Recall (0.91): The model is effective at identifying the positive class ("Yes").

- Moderate Precision (0.71): There are some False Positives, reducing precision.

- F1 Score (0.80): Balances precision and recall, indicating overall good performance.

- Accuracy (0.85): 85% of all predictions are correct.


### 9. Random Forest Model

A Random Forest model is trained and evaluated
 
```{r}
#Fitting Random Forest Using randomForest package
library(randomForest)
data_n$Target <- factor(data$Target)
library(caTools)
set.seed(101)
split = sample.split(data_n$Target, SplitRatio = 0.70)
train = subset(data_n, split == TRUE)
test = subset(data_n, split == FALSE)
model <- randomForest(Target ~ Age + Sex + ChestPain + RestBP + Chol + Fbs  + RestECG + MaxHR + ExAng + Oldpeak + Slope + Ca + Thal,data=train)
rf.pred <- predict(model,test)
table(rf.pred,test$Target)

```

```{r}
# Define confusion matrix values
TP <- 33  # True Positives
TN <- 40  # True Negatives
FP <- 9   # False Positives
FN <- 9   # False Negatives

# Accuracy
accuracy <- (TP + TN) / (TP + TN + FP + FN)
print(paste("Accuracy:", round(accuracy, 4)))

# Precision for class 1
precision_1 <- TP / (TP + FP)
print(paste("Precision (class 1):", round(precision_1, 4)))

# Recall for class 1
recall_1 <- TP / (TP + FN)
print(paste("Recall (class 1):", round(recall_1, 4)))

# F1-score for class 1
f1_score_1 <- 2 * (precision_1 * recall_1) / (precision_1 + recall_1)
print(paste("F1-Score (class 1):", round(f1_score_1, 4)))

# Precision for class 0
precision_0 <- TN / (TN + FN)
print(paste("Precision (class 0):", round(precision_0, 4)))

# Recall for class 0
recall_0 <- TN / (TN + FP)
print(paste("Recall (class 0):", round(recall_0, 4)))

# F1-score for class 0
f1_score_0 <- 2 * (precision_0 * recall_0) / (precision_0 + recall_0)
print(paste("F1-Score (class 0):", round(f1_score_0, 4)))

```

#### Interpretation:

- Accuracy (80.22%) - The model correctly classified 80.22% of all cases, showing reliable overall performance in distinguishing between positive and negative classes.

- Precision for Class 1 (78.57%) - When predicting the positive class, the model is correct 78.57% of the time, minimizing false positives effectively.

- Recall for Class 1 (78.57%) - The model identifies 78.57% of actual positive cases, indicating a moderate ability to avoid missing true positives.

- F1-Score for Class 1 (78.57%) - A balanced measure of precision and recall, showing the model is reliable in predicting the positive class.

- Precision for Class 0 (81.63%) - When predicting the negative class, the model is correct 81.63% of the time, effectively avoiding false negatives.

- Recall for Class 0 (81.63%) - The model correctly identifies 81.63% of actual negative cases, indicating robustness in detecting the negative class.

- F1-Score for Class 0 (81.63%) - A balanced measure of precision and recall for the negative class, demonstrating consistent reliability


### 10. Model Comparison

A confusion matrix and metrics are presented for all models.

```{r}
# Create the table as a data frame
model_performance <- data.frame(
  Model = c("Logistic Regression", "Support Vector Machine", "Random Forest"),
  Accuracy = c("77.05%", "83.61%", "80.22%"),
  Precision = c("81.82%", "71.43%", "78.57%"),
  Recall = c("64.29%", "90.91%", "78.57%"),
  F1_Score = c("72.00%", "80.00%", "78.57%"),
  Specificity = c("87.88%", "N/A", "81.63%")
)

# Generate the table
model_performance %>%
  gt() %>%
  tab_header(title = "Model Performance Comparison") %>%
  cols_label(
    Model = "Model",
    Accuracy = "Accuracy",
    Precision = "Precision",
    Recall = "Recall",
    F1_Score = "F1 Score",
    Specificity = "Specificity"
  ) %>%
  fmt_percent(columns = starts_with("Accuracy"), decimals = 2)

```

### 11. Conclusion

SVM achieved the highest accuracy and recall, making it effective for identifying positive cases.
Logistic Regression excelled in precision and specificity but struggled with recall.
Random Forest provided balanced metrics across all measures, making it a reliable choice.
For high recall requirements, SVM is the best model.
Random Forest is recommended for balanced performance in diverse applications.



### 12. Insights:

This project demonstrates the effectiveness of predictive modeling for heart disease using Logistic Regression, Support Vector Machine (SVM), and Random Forest. Among the models, SVM achieved the highest recall (90.91%), making it ideal for scenarios prioritizing the identification of true positives, such as early disease detection. However, Random Forest provided balanced performance across metrics, suggesting it is better suited for applications requiring a balance between sensitivity and specificity.